An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2010-12-09 paper.tex
Working on Section 2...

\usepackage{amsmath}
\usepackage{hyperref}

@@

\begin{abstract}
  We do something.
\end{abstract}

@@

>>>>

\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{color}

\newcommand{\vtheta}{\vec{\theta}}

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\bel}[1]{\begin{equation}\label{#1}}
\newcommand{\ba}{\begin{eqnarray}}
\newcommand{\ea}{\end{eqnarray}}
\newcommand{\bal}[1]{\begin{eqnarray}\label{#1}}

\def\ilya{\textcolor{red}}
\def\will{\textcolor{blue}}

\bibliographystyle{h-physrev}

@@

\affiliation{MIT Kavli Institute; and University of Birmingham, UK}

\begin{abstract}
\ilya{Add physics/astro motivation}
  Reversible-jump Markov chain monte carlo (RJMCMC) is an extremely
  powerful technique for performing Bayesian model selection, but it
  suffers from a fundamental difficulty: it requires jumps between
  model parameter spaces, but cannot retain a memory of the favored
  locations in more than one parameter space at a time.  Thus, a naive
  jump between parameter spaces is unlikely to be accepted in the MCMC
  algorithm and convergence is correspondingly slow.  Here we
  demonstrate an interpolation technique that uses samples from
  single-model MCMCs to propose inter-model jumps from an
  approximation to the single-model posterior of the target parameter
  space.  The interpolation technique, based on a kD-tree
  data structure, is adaptive and efficient in arbitrary dimensions.
  We show that our technique leads to dramatically improved convergence
  over naive jumps in an RJMCMC, and compare it to other proposals in
  the literature to improve the convergence of RJMCMCs.  We also
  discuss the use of the same interpolation technique in two other
  contexts: as a convergence test for a single-model MCMC and as a way
  to construct efficient ``global'' proposal distributions for
  single-model MCMCs without prior knowledge of the structure of the
  posterior distribution.
\end{abstract}

@@

\section{Introduction}

\section{Reversible Jump MCMC}

Reversible jump MCMC (RJMCMC) \cite{Green1995} is a technique for Bayesian model comparison.  Below, we give a very brief introduction to Bayesian analysis; describe a standard MCMC; and introduce RJMCMC.

\subsection{Bayesian analysis}

Consider an observed data set $d$ and a set of competing models for the data, indexed by an integer $i$: $\{M_i | i = 1, 2, \ldots \}$.  Each model has some continuous parameters, $\vtheta_i$; given the model and its parameters, we can make a prediction about the likelihood of observing
the experimental data: $L(d|\vtheta_i, M_i)$.  Within the framework of each model, Bayes' rule gives us a way to compute the posterior probability distribution function (PDF) for the model
parameters implied by the data:
\begin{equation}
  p(\vtheta_i | d, M_i) = \frac{L(d|\vtheta_i, M_i) p(\vtheta_i|M_i)}{p(d|M_i)},
\end{equation}
where $p(M_i, \vtheta_i |d)$ is the posterior distribution for the
model parameters $\vtheta_i$ implied by the data in the context of
model $M_i$, $p(\vtheta_i|M_i)$ is the prior
probability of the model parameters that 
represents our beliefs before accumulating any of the data $d$, and
$p(d)$ is an overall normalizing constant that ensures that $p(\vtheta_i|d,M_i)$ is properly normalized as a probability distribution on
the $\vtheta_i$.  This implies that the evidence is equal to
\bel{evidence}
  p(d|M_i) = \int_{V_i} d\vtheta_i L(d|\vtheta_i, M_i) p(\vtheta_i|M_i),
\end{equation}
where $V_i$ is the parameter space volume in model $M_i$.  We are interested in the evidence for model $M_i$ given the data, $p(M_i|d)$, which is again given by Bayes' rule as
\begin{equation}
p(M_i|d) = \frac{p(d|M_i) p(M_i)}{p(d)},
\end{equation},
where $p(M_i)$ is our a priori belief in model $M_i$ and $p(d)$ is a normalizing constant, 
\be
p(d)=\sum_i p(d|M_i) p(M_i).
\ee

When selecting among alternative models, we are interested in finding the model with the highest evidence $p(M_i|d)$.  However, attempts to directly compute the evidence by performing the integration in Eq.~(\ref{bel}) are generally very difficult in a multi-dimensional, multi-modal parameter space when the likelihood has to be evaluated numerically.  In particular, a grid-based integral quickly becomes computationally unfeasible as the dimensionality of $\vtheta$ exceeds a few.  The parameter space must typically be explored in a stochastic manner before the evidence integral can be computed.  Although several stochastic parameter-exploration techniques are focused directly on evidence computation (e.g., nested sampling and MultiNest \ilya {Add references}), we frequently want to compute the posterior PDFs within each model along with the evidences for the various models, and the Markov chain Monte Carlo is one of the most widely used tools for this purpose.

\subsection{MCMC}

MCMC methods produce a Markov chain (or sequence) of parameter
samples, $\{ \vtheta^(i) \, | \, i = 1, \ldots \}$, such that a particular
parameter set, $\vtheta$, appears in the sequence with a frequency
equal to its probability according to a posterior, $p(\vtheta|d)$.  A
Markov chain has the property that the transition probability from one
element to the next, $p(\vtheta_i \to \vtheta_{i+1})$, depends only on
the value of $\vtheta_i$, not on any previous values in the chain.

One way to produce a sequence of MCMC samples is via the following
algorithm, first proposed by \citet{Metropolis1953} and used widely in
the physical sciences thereafter:
\begin{enumerate}
  \item Begin with the current sample, $\vtheta_i$.
  \item Propose a new sample, $\vtheta_p$, by drawing randomly from a
    ``jump proposal distribution'' with probability $Q(\vtheta_i \to
    \vtheta_p)$.  Note that $Q(\vtheta_i \to \vtheta_p)$ can depend on
    the current parameters, $\vtheta_i$, and any other ``constant''
    data, but cannot examine the history of the chain beyond the most
    recent point.  This is necessary to preserve the Markovian
    property of the chain.
  \item Compute the ``acceptance'' probability,
    \begin{equation}
      \label{eq:paccept}
      p_{\textnormal{accept}} \equiv
      \frac{p(\vtheta_p|d)}{p(\vtheta_i|d)} \frac{Q(\vtheta_p \to
        \vtheta_i)}{Q(\vtheta_i \to \vtheta_p)}
    \end{equation}
  \item With probability $\min(1,p_{\textnormal{accept}})$ ``accept''
    the proposed $\vtheta_p$, setting $\vtheta_{i+1} = \vtheta_p$;
    otherwise set $\vtheta_{i+1} = \vtheta_i$.
\end{enumerate}
This algorithm is more likely to accept a proposed jump when it
increases the posterior (the first factor in Equation
\eqref{eq:paccept}) and when it is to a location in parameter space
from which it is easy to return (the second factor in Equation
\eqref{eq:paccept}); the combination of these influences in Equation
\eqref{eq:paccept} ensures that the equilibrium distribution of the
chain is $p(\vtheta|d)$.  As $i \to \infty$ the samples $\vtheta_i$
are distributed according to $p(\vtheta|d)$.

In practice the number of samples required before the chain
appropriately samples $p(\vtheta|d)$ depends strongly on the jump
proposal distribution; proposal distributions that often propose jumps
toward or within regions of large $p(\vtheta|d)$ can be very
efficient, while poor proposal distributions can require prohibitively
large numbers of samples before convergence.

There is no foolproof test for the convergence of a chain.  In this
work, we test the convergence of our chains by comparing the
statistics calculated from the entire chain to statistics calculated
from only the first half of the chain; when the chain has converged,
the two calculations agree.  This is a necessary, but not sufficient,
condition for convergence.

We begin the chain at an arbitrary point in parameter space; this is
equivalent to taking a finite section of an infinite chain that begins
with the chosen point.  Every point in parameter space occurs in an
infinite chain, and no section of the chain is better than any other,
so a sufficiently long, but finite, section of the infinite chain
chosen in this manner can be representative of the statistics of the
chain as a whole.  However, because consecutive samples in a chain are
correlated with each other, the beginning of our finite chain has a
``memory'' of the starting point; we discard enough points at the
beginning of the finite chain that we can be confident it does not
retain a memory of the arbitrary starting point.  The points discarded
in this way are commonly called ``burn-in'' points.

\subsection{RJMCMC}

Consider the problem of model selection among a set of models, and the
``super-model'' that encompasses all the models under consideration.
The parameter space of the super-model consists of a discrete
parameter that identifies the choice of model, $M_i$, and the
continuous parameters appropriate for this model, $\vtheta_i$.  We
denote a point in the super-model parameter space by $\{M_i,
\vtheta_i\}$; each such point is a statement that, e.g., ``the
underlying mass distribution for black holes in the galaxy is a
Gaussian, with parameters $\mu$ and $\sigma$,'' or ``the underlying
mass distribution for black holes in the galaxy is a triple-bin
histogram with parameters $w_1$, $w_2$, $w_3$, and $w_4$,'' or ....
To compare models, we are interested in the quantity (see Equation
\eqref{eq:model-posterior-def})
\begin{equation}
  p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d).
\end{equation}
If we perform an MCMC in the super-model parameter space, then we
obtain a chain of samples $\{M_i, \vtheta_i \, | \, i = 1, \ldots\}$
distributed in parameter space with density $p(M_i,\vtheta_i|d)
d\vtheta_i$ and we can estimate the integral as
\begin{equation}
    p(M_i|d) = \int d\vtheta_i p(M_i, \vtheta_i|d) \approx \frac{N_i}{N},
\end{equation}
where $N_i$ is the number of samples in the chain lying in the
parameter space of model $M_i$ and $N$ is the total number of samples
in the chain.  The fraction of samples lying in the parameter space of
model $M_i$ gives the probability of that model relative to the other
models under consideration.

To perform the MCMC in the super-model parameter space, we must
propose jumps not only between points in a particular model's
parameter space, but also between the parameter spaces of different
models.  For this MCMC to be efficient, proposed jumps into a model
from another should favor regions with large posterior; when the
posterior is highly-peaked in a small region of parameter space,
proposed jumps outside this region are unlikely to be accepted, and
the reversible-jump MCMC samples will require a very long chain to
properly sample the ``super-model'' posterior.

We can exploit the information we have from single-model MCMCs to
generate efficient jump proposal distributions for our reversible jump
MCMC.  We would like to propose jumps that roughly follow the
distribution of samples in the single-model MCMCs.

\section{kD Trees and Interpolation}

Mention example of bricks from \cite{Littenberg2009}.

\section{RJMCMC Efficiency}

\section{Example}

BH mass paper [still not sure how to present this without repetition...  ]

\section{Conclusion/Discussion}

...

\subsection{Other Uses}
The efficient interpolation of a PDF via a kD-tree described in this paper can be extremely useful in other contexts, beyond a reversible-jump MCMC.  We suggest two examples below: generating efficient jump proposal distributions and convergence tests?

...

Mention nested sampling?

\nocite{Littenberg2009}

\bibliography{paper}
